Step by Step: Using Step Selection Analysis to Simulate Dispersal and Assess Landscape Connectivity

### 2021-05-05 Chapter_1/04_Manuscript/Manuscript.tex
Various updates

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
% \usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}

@@

\textsuperscript{2} Botswana Predator Conservation Trust, Private Bag 13, Maun,
Botswana.

@@

\textbf{Keywords:} dispersal, simulation, integrated step selection,
Kavango-Zambezi Transfrontier Conservation Area, landscape connectivity, Lycaon
pictus

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
For many species, dispersal represents a crucial process that allows to avoid
inbreeding, to colonize new habitats and to reinforce small and unviable
subpopulations. Successful dispersal is therefore a crucial pre-requisite for
long-term species persistence in wild animal populations. However, the ability
to disperse is contingent a sufficient degree of landscape connectivity, which
is why the estimation of connectivity and identification of dispersal corridors
has become a task of extraordinary importance in ecological studies and for the
planning of conservation initiatives.

Over the past two decades, scientists have mainly relied on analytical tools
such as least-cost methods and circuit theory to investigate landscape
connectivity. This is largely owed to their ease of use and intuitive nature,
which have facilitated their application in the past. Despite their usefulness
for various ecological applications, both methods make several restricting
assumptions that can limit their applicability in reality. Least-cost analysis,
for instance, suffers from the fact that animals are assumed to have perfect
knowledge about the entire landscape and choose cost-minimizing routes
accordingly. On the contrary, circuit theory assumes that animals walk according
to a complete random walk, which is equally unlikely to hold in reality.

Fortunately, recent innovations in movement ecology have brought forward novel
opportunities to study dispersal and landscape connectivity. In particular, the
rich suite of resource selection functions has undergone substantial
improvements in recent years. Most notably, step selection functions have been
generalized to \textit{integrated} step selection functions, which essentially
represent fully mechanistic movement models based on which individual animal
movement could be predicted and simulated. While such models have been applied
to simualte the steady state utilization distribution, similar approaches could
be useful for investigating transient movement behavior and study landscape
connectivity in a much more natural way.

Here, we showcase the use of integrated step selection analysis to simulate
dispersal of a top-predator across a large study area. For this, we collected
data on 16 dispersing wild dogs and analysed it in an integrated step selection
analysis to produce a fully mechanistic movement model rendering wild dog
dispersal. Based on this model, we simulated 80'000 dispersers moving across the
Kavango-Zambezi Transfrontier Conservation Area, the world's largest
transboundary conservation area. We summarize the simulation output in a heatmap
indicating the intensity of use across the KAZA-TFCA, and we apply network
theory to produce a set of complementary maps that further reveal major
dispersal hotspots and crucial bottlenecks. Finally, we discuss the benefits and
pitfalls of dispersal simualtions and we highlight improvements to be made in
the future.
\end{abstract}

@@

% % Create linenumbers
% \linenumbers

\section{Introduction}
% Importance of Dispersal & Connectivity
Dispersal of individuals is an important process governing the dynamics wild
animal populations that are distributed in space \citep{Clobert.2012}. It is
defined as the movement of individuals from their natal location to the site of
first reproduction \cite{Howard.1960} and allows species to avoid inbreeding
\citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012}, to rescue small and
unviable populations \citep{Brown.1977}, and to promote the colonization of
unoccupied habitats \citep{Hanski.1998, MacArthur.2001}. Successfull dispersal
requires a sufficient degree of landscape connectivity, which is why the
protection of dispersal corridors has become a major goal in conservation
science. Information on movement behavior during dispersal and knowledge about
the factors that limit dispersal is therefore critical for a comprehensive
understanding of landscape connectivity and population viability
\citep{Baguette.2013, Vasudev.2015}. In addition, reliable modelling techniques
to identify dispersal corridors based on empirical data are necessary
\citep{Diniz.2019}.

% Advancements in GPS Technology and Movement Ecology
Thanks to novel technologies developed over the past decades, particularly of
GPS/Satellite radio-collars, the study of dispersal and connectivity using
telemetry data has accelerated \citep{Jonsson.2016, Williams.2019}.
Additionally, the advent of publicly accessible satellite imagery and
sophisticated remote sensing techniques to represent the physical landscape
through which individuals disperse have heralded the ``golden age of animal
tracking'' \citep{Kays.2015}. Concurrently, the increased availability of large
amounts of empirical data and an increased computational power have led to the
development of several modelling techniques that allow studying movement
behavior during dispersal and highlighting major movement corridors
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Zeller.2012, Diniz.2019}.

% Resource Selection & Connectivity
\textit{Resource selection functions} \citep{Boyce.2002} and derived methods
such as \textit{step selection functions} \citep{Fortin.2005} and \textit{path
selection functions} \citep{Cushman.2010} have proven particularly useful for
studying animal movement \citep{Fieberg.2020}. These methods allow estimating
habitat preferences of the focal species by comparing covariates at locations
visited by the animal to the scame covariates at locations available to, but not
visited by the animal. The so estimated preferences can then be used to predict
a permeability surface, indicating the expected ease at which an animal can
traverse a given area \citep{Zeller.2012}. Utimately, the permeability surface
serves as input to a connectivity model that is used to reveal movement
corridors. In this regard, two of the most prominent connectivity models in the
literature are least-cost path analysis \citep{Adriaensen.2003} and circuit
theory \citep{McRae.2006, McRae.2008}.

% Least-cost path
In least-cost path analysis, connectivity is estimated by identifying
least-costly routes between pre-defined start- and endpoints
\citep{Adriaensen.2003}. For this, the permeability surface is converted into a
network graph and permeability scores are translated into probabilities for
moving from one cell to another. Least-costly routes between start- and
endpoints can then be computed using Dijkstra’s algorithm \citep{Dijkstra.1959}.
Although intuitively appealing, the method suffers from several weaknesses.
First of all, a least-cost path always exist, even if the associated costs are
unreasonably high. Second, it assumes that animals have a preconceived end-point
in mind and choose a cost-minimizing route accordingly. While this assumption
may be reasonable for migrating animals that move between a limited number of
habitats, it is unlikely to hold for dispersers that move over long distances
into unknown territory \citep{Koen.2014, Abrahms.2017, Cozzi.2020}. Finally, the
method also requires that animals have an infinite perceptual range, otherwise
they could not compute an optimal path. Some of these issues have been addressed
using alternative least-cost approaches. For instance, instead of computing a
least-cost path, a least cost-corridor that also considers slightly suboptimal
routes \citep{Pinto.2009} can be calculated. The corridor relaxes the assumption
of perfect knowledge and overcomes the single pixel width issue. Alternatively,
the randomized least-cost path has been proposed, which allows animals to
deviate from the least-cost route by a pre-defined factor \(\theta\)
\citep{Panzacchi.2016, VanMoorter.2021}.

% Circuit theory
In contrast to least-cost approaches, circuit theory makes use of electrical
circuit theory to quantify landscape connectivity \citep{McRae.2006,
McRae.2008}. Here, the permeability surface is converted into an electrical
circuit containing nodes and resistor, where resistance values depend on habitat
permeability. A source point is connected to a current source, whereas an
endpoint is grounded. Hence, the current flowing through each node in the
circuit can be computed and is supposed to resemble the relative frequency at
which each location in the study area is being visited. By revealing not only a
single path or corridor, circuit theory overcomes the single-pixel width issue
inherent to least-cost approaches. Intrinsically, circuit theory is based on the
assumption that animals follow a random walk and can therefore not render
directional biases. In reality, however, dispersers often move in a very
directional fashion as they want to cover as much ground in as little time as
possible. Furthermore, it is assumed that individuals have a perceptual range of
a single pixel, which rarely captures the true spatial scale of selection (cite
someone).

% Temporal scale
Neither least-cost analyes or circuit theory are capable of rendering the
temporal dimension of movement. Such influences could, however, be studied using
individual based movement models. Still, only few have explored these
possibilities \citep{Kanagaraj.2013, Hauenstein.2019, Zeller.2020}.

% Step selection analysis
While step selection functions were initially intended to learn about habitat
preferences of the focal species \citep{Fortin.2005}, recent improvements may
elevate its applicability into other areas. Originally, \cite{Fortin.2005}
proposed step selection functions to account for serial autocorrelation inherent
to GPS relocation data. In this method, covariates along realized steps (the
connecting lines between two consecutive relocations; ; \citealp{Turchin.1998})
are contrasted to covariates along  ``alternative''  or  ``random steps'' that
are generated by randomly sampling turning angles and step lengths based on
observed movement characteristics. It is then assumed that animals assign a
selection score \(w(x)\) of the following form to each step:

@@

The selection score \(w(x)\) of each step depends on its associated covariates
(\(x_1, x_2, ..., x_n\)) and on the animal's relative selection strengths (i.e.
preferences) towards these covariates (\(\beta_1, \beta_2, ..., \beta_n\)). The
probability that a step \(i\) is realized \(P(Y_{i} = 1\)) then depends on the
step's selection score, as well as on the selection scores of all alternative
steps:

@@

\noindent Preferences of interest, i.e. the \(\beta\)'s, can be estimated by
comparing realized (scored 1) and random (scored 0) steps in a conditional
logistic regression model \citep{Fortin.2005}. In this model, positive
\(\beta\)-coefficients indicate selection of a covariate, whereas negative
\(\beta\)-coefficients indicate avoidance of a covariate. To deal with multiple
individuals, one can either average estimates of individual models
\citep{Murtaugh.2007, Fieberg.2010} or apply mixed effects conditional logistic
regression analysis as recently proposed by \cite{Muff.2020}.

% Integrated step selection analysis
An animals movement trajectory can be viewed as the result of an interplay
between habitat and movement preferences. Traditional step selection functions
cannot account for a correlation between turning angles and step lengths, unless
the two are sampled jointly from a copula distribution \citep{Hodel.2021}.
Recently, however, SSFs have been generalized to \textit{integrated} SSFs
(iSSFs), which tear apart movement and habitat preferences of the studied
animals \citep{Avgar.2016}. The method thus allows to render an animals
preferences with respect to prevailing habitat conditions, as well as with
respect to movement characteristics. This reduces potential biases in estimated
preferences and enables parametrization of a fully mechanistic movement model
from which movement and space use can be simulated \citep{Avgar.2016,
Signer.2017}. \cite{Signer.2017} used integrated step selection analysis to
simulate steady state utilization distributions of resident animals. However,
the degree to which such simulations could be used to simulate dispersers and
infer movement corridors is unknown.

% Study Species & Study Area
One of the species for which long-term species persistence relies on sufficient
landscape connectivity is the endangered African wild dog \textit{Lycon pictus}.
This species has once been wide-spread across sub-Saharan Africa but disappeared
from a vast majority of its historic range due to ongoing persecution by humans,
habitat destruction, and deadly diseases. As of today, only 6'000 free-ranging
individuals remain in small and spatially scattered subpopulations. Within those
subpopulations, wild dogs form cohesive packs comprising 8 to 12 adults and
their offspring \cite{McNutt.1995}. After reaching sexual maturity, male and
female offspring form same-sex coalitions and disperse from their natal pack
\citep{McNutt.1996, Behr.2020}. New packs are formed when dispersing coalitions
join unrelated opposite-sex dispersing coalitions \citep{McNutt.1996}.
Dispersing wild dogs can cover several hundred kilometers across a variety of
landscapes \citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020, Hofmann.2021}.
One of the few strongholds for this species lies near the Moremi Game Reserve in
northern Botswana, which is part of the world's largest transboundary protected
area, namely the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA).
This area has originally been intended to facilitate migration of elephants, but
is expected to benefit a multitude of other species \citep{Elliot.2014,
Brennan.2020, Hofmann.2021}.

In a previous paper, we assessed landscape connectivity within the KAZA-TFCA for
dispersing African wild dogs using a least-cost corridor approach. For this, we
fitted a basic habitat selection model based on which we predicted landscape
permeability. We now expand on this knowledge and develop a more detailed
movement model of dispersing wild dogs. We then use this model to simulate
thousands of dispersers moving across the KAZA-TFCA. Based on said simulations,
we compute heatmaps and identify potential dispersal hotspots and compare them
to the dispersal routes identified in \citep{Hofmann.2021}. We also showcase how
simulated dispersal data can be analysed using network-analysis and how network
metrics relevant to landscape connectivity can be computed. Our results show
that a simulation based approach yields several major benefits over traditional
connectivity modelling techniques. Most importantly, simulations provide a more
generic view on how connectivity emerges and to which degree connectivity
depends on the dispersal duration. In addtion, by generating proper dispersal
trajectories, network theory can be applied to calculate network metrics that
are pertinent to connectivity analysis. Finally, we put forward additional
opportunities using simulations that go beyond the scope of this paper.

>>>>

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}

@@

\textsuperscript{2} Botswana Predator Conservation, Private Bag 13, Maun,
Botswana.

@@

\textbf{Keywords:} dispersal, simulation, movement, integrated step selection,
Kavango-Zambezi Transfrontier Conservation Area, landscape connectivity, Lycaon
pictus

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Dispersal is an important process that allows species to avoid inbreeding, to
colonize new habitats and to reinforce non-viable subpopulations. Successful
dispersal thus represents a crucial pre-requisite for long-term species
persistence in wild animal populations. However, the ability to disperse is
contingent a sufficient degree of landscape connectivity, which is why the
estimation of connectivity and identification of dispersal corridors has become
a task of extraordinary importance in ecological studies.

Over the past two decades, ecologists have primarily relied on analytical tools
such as least-cost methods and circuit theory to investigate landscape
connectivity. Despite their usefulness for various ecological applications, both
methods make several restricting assumptions that limit their suitability in
reality. To address their shortcomings, simulations from individual-based
movement models have been proposed and applied. To overcome their shortcomings,
simulations from individual-based movement models have been proposed. Yet, due
to the infinite number of possibilities of coming up with an individual-based
model, a unified and objective framework is missing.

Recent innovations in movement ecology have brought forward novel opportunities
to study animal dispersal and landscape connectivity. In particular, the rich
suite of resource selection functions, including point-, step-, and
path-selection functions have undergone substantial improvements over the past
years. Most notably, step-selection functions have been generalized to
\textit{integrated} step selection functions, which essentially represent fully
mechanistic movement models based on which individual animal movement could be
simulated. While such models have been applied to study \textit{steady-state}
utilization distribution, a similar approach may be useful for investigating
\textit{transient} movement behavior and thereby study landscape connectivity by
means of simulated dispersal events.

Here, we showcase the use of integrated step selection analysis to simulate
dispersal of the highly endangered African wild dog across the world's largest
transboundary conservation area, the Kavango-Zambezi Transfrontier Conservation
Area (KAZA-TFCA). For this, we utilize data collected on 16 dispersing wild dog
coalitions departing from northern Botswana. We combine the movement data with
relevant spatial covariate layers and used integrated step selection functions
to parametrize a fully mechanistic movement model rendering wild dog dispersal.
Based on this model, we simulate 80'000 dispersers, originating from protected
areas and moving across the extent of the KAZA-TFCA. We then generate a heatmap
and use network theory to reveal dispersal hotspots and crucial bottlenecks
across the study area. Finally, we discuss the benefits and pitfalls of such
dispersal simualtions and highlight potential improvements to be made in the
future.
\end{abstract}

@@

% Create linenumbers
\linenumbers

\section{Introduction (90\%)}

% Importance of Dispersal & Connectivity
\subsection{Importance of Dispersal \& Connectivity (90\%)}
Dispersal is defined as the movement of individuals from their natal location to
the site of first reproduction \cite{Howard.1960}. It is a vital process
governing the dynamics wild animal populations that are distributed in space
\citep{Hanski.1998, Clobert.2012} and may strongly affect population dynamics at
different spatial and social scales \citep{Hanski.1999a, Clobert.2012}.
Dispersal allows species to avoid inbreeding and maintain genetic diversity
\citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012, Baguette.2013}, to
rescue small and non-viable populations \citep{Brown.1977}, and to promote the
colonization of unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}.
However, successful dispersal requires a sufficient degree of functional
connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the identification
and protection of major dispersal corridors has become an important task in
conservation science \citep{Doerr.2011, Rudnick.2012}. To achieve this, reliable
information on movement behavior during dispersal and knowledge about factors
that limit dispersal and connectivity is paramount \citep{Baguette.2013,
Vasudev.2015}.

% Advancements in GPS Technology & Movement Ecology
\subsection{Advancements in GPS Technology \& Movement Ecology (90\%)}
Thanks to novel technologies developed over the past decades, particularly of
GPS/Satellite radio-collars, the use of GPS data to study animal dispersal and
connectivity has accelerated \citep{Elliot.2014, Jonsson.2016, Williams.2019}.
Additionally, the advent of publicly accessible satellite imagery and
sophisticated remote sensing techniques to represent the physical landscape
through which individuals disperse have heralded a ``golden age of animal
tracking'' \citep{Kays.2015}. Concurrently, the availability of large amounts of
empirical data and an increased computational power have led to the development
of numerous techniques to study dispersal movements and highlight critical
corridors between subpopulations \citep{Boyce.2002, Fortin.2005, Cushman.2010,
Zeller.2012, Diniz.2020}.

% Resource Selection & Connectivity
\subsection{Resource Selection \& Connectivity (90\%)}
\textit{Resource selection functions} \citep{Boyce.2002} and derived methods
such as \textit{step selection functions} \citep{Fortin.2005} and \textit{path
selection functions} \citep{Cushman.2010} have proven particularly useful for
studying animal movement \citep{Fieberg.2020} and modelling connectivity
\citep{Diniz.2020}. These methods allow estimating habitat preferences of the
focal species by comparing covariates at locations visited by the animal to the
same covariates at locations available to, but not visited by the animal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Thurfjell.2014}. The so estimated
preferences can then be used to predict a permeability surface, indicating the
expected ease at which an animal can traverse a given area \citep{Spear.2010,
Zeller.2012, Etherington.2016}. Ultimately, the permeability surface serves as
input to a connectivity model that is used to reveal movement corridors
\citep{Diniz.2020}. Two of the most prominent connectivity models are least-cost
path analysis (LCP analysis; \citealp{Adriaensen.2003}) and circuit theory (CT
\citealp{McRae.2006, McRae.2008}), both graph-based methods that estimate
conductance of the landscape. Despite their intuitive nature and ease of use,
both methods make rigorous assumptions about animal movement that may or may not
be fulfilled in reality \citep{Diniz.2020}.

% Issues with Least-Cost Paths & Circuit Theory
\subsection{Issues with Least-Cost Paths \& Circuit Theory (90\%)}
In LCP analysis, for instance, a least costly path always exists, even if
associated movement costs are unreasonably high and will never be incurred by a
dispersing individual. The method also presumes that animals have an infinite
perceptual range, a preconceived end-point in mind, and choose a cost-minimizing
route accordingly. These assumptions may be reasonable for migrating animals,
yet they are unlikely to be true for dispersers that move over long distances
into unknown territory \citep{Koen.2014, Abrahms.2017, Cozzi.2020}. Finally,
LCPs are only one pixel wide, meaning that their absolute size depends on the
resolution of chosen covariate layers \citep{Diniz.2020}. Although some of these
issues can be alleviated using less stringent versions of the LCP algorithm
(e.g. least-cost \textit{corridors} \citep{Pinto.2009}, \textit{thresholded}
least-cost paths \citep{Landguth.2012}, and \textit{randomized} least-cost paths
\citep{Panzacchi.2016, VanMoorter.2021}), a certain degree of arbitrariness in
the assumptions remains. CT entails similar restrictions. Because CT only
considers movements from the source cell to its 4 or 8 adjacent cells, it
implicitly posits a perceptual range of a single pixel. It is also built on the
assumption of a complete random walk \citep{Diniz.2020}, implying that
directional biases cannot be rendered, albeit being very common in dispersal
movements \citep{Cozzi.2020, Hofmann.2021}. Ultimately, both LCP analysis and CT
omit the temporal dimension of dispersal. In result, statements about the
expected duration required to traverse a certain corridor are impossible.
Because movement is not modelled explicitely, neither of the methods can render
possible interactions between movement behavior and landscape characteristics
such that connectivity mainly arises as a result of the landscape structure.
This is referred to as structural connectivity and stands in contrast to
functional connectivity, which also renders the behavioral response of the
animal with respect to prevailing habitat conditions \citep{Tischendorf.2000}.
Even though functional connectivity is more difficult to estimate, a functional
view is the ultimate goal in conservation science because it has direct
consequences for gene flow \citep{Baguette.2013}.

% What about IBMMs?
\subsection{What about IBMMs? (90\%)}
To address the issues inherent to LCPs and CT, individual-based movement models
(IBMMs) have been proposed and applied \citep{Diniz.2020}. In these models,
dispersal is simulated explicitly, based on movement rules that determine how
individuals move over and interact with the prevailing landscape
\citep{Kanagaraj.2013, Clark.2015, Allen.2016, Hauenstein.2019, Zeller.2020,
Vasudev.2021}. Using the simulated trajectories, one can calculate a set of
connectivity metrics, such as interpatch-connectivity and traversal frequency
across the landscape to reveal major dispersal hotspots \citep{Kanagaraj.2013,
BastilleRousseau.2018, Hauenstein.2019, Zeller.2020}. Even though IBMMS can be
employed to overcome any of the shortcomings intrinsic to LCPs and CT, as well
as to provide a more functional view on connectivity, they can be challenging to
fit and require vast amounts of data collected during dispersal
\citep{Diniz.2020}.

% Step Selection Analysis
\subsection{Step Selection Analysis (90\%)}
Here, we investigate the usefulness of integrated step selection functions
(ISSFs, \citealp{Avgar.2016}), as a relatively simple but powerful IBMM based on
which dispersal can be simulated. While regular SSFs were intended to learn
about relative habitat preferences of the focal species \citep{Fortin.2005,
Thurfjell.2014, Avgar.2017}, the method has been generalized and now enables to
jointly study habitat and movement preferences, as well as potential
interactions between movement and habitat preferences \citep{Avgar.2016,
Signer.2017, Fieberg.2020}. ISSFs therefore provide a relatively simple method
to model complex movement behavior where movement results from two intertwined
behavioral kernels (e.g. \citealp{Prokopenko.2017, Munden.2020}). Importantly, a
parametrized ISSF model can be viewed as a fully mechanistic movement model
based on which individual movement trajectories can be simulated
\citep{Avgar.2016, Signer.2017}. In fact, \cite{Signer.2017} used ISSF to
simulate steady state utilization distributions of resident animals. However,
the degree to which such simulations are helpful in detecting movement corridors
and modeling landscape connectivity is unknown.

% Study Species & Study Area
\subsection{Study Species \& Study Area (90\%)}
One of the species for which long-term viability relies on sufficient
landscape connectivity is the endangered African wild dog \textit{Lycon pictus}.
While once present across entire sub-Saharan Africa, wild dogs have disappeared
from a vast majority of their historic range due to persecution by humans,
habitat destruction, and deadly diseases. As of today, only 6'000 free-ranging
individuals remain in small and spatially scattered subpopulations
\citep{Woodroffe.2012}. Within those subpopulations, wild dogs form cohesive
packs comprising 8 to 12 adults and their offspring \cite{McNutt.1995}. After
reaching sexual maturity, male and female offspring form same-sex coalitions and
disperse from their natal pack \citep{McNutt.1996, Behr.2020}. New packs are
formed when dispersing coalitions join unrelated opposite-sex dispersing
coalitions \citep{McNutt.1996}. Dispersing wild dogs can cover several hundred
kilometers across a variety of landscapes \citep{DaviesMostert.2012,
Masenga.2016, Cozzi.2020, Hofmann.2021}. One of the few strongholds for this
species lies near the Moremi Game Reserve in northern Botswana, which is part of
the world's largest transboundary protected area, namely the Kavango-Zambezi
Transfrontier Conservation Area (KAZA-TFCA). This area has originally been
intended to facilitate migration of elephants, but is expected to benefit a
multitude of other species \citep{Elliot.2014, Brennan.2020, Hofmann.2021}.

% Previous Paper
\subsection{Previous Paper (90\%)}
In a previous study, we assessed landscape connectivity for dispersing African
wild dogs within the KAZA-TFCA using a least-cost corridor approach
\citep{Hofmann.2021}. For this, we fitted a basic habitat selection model based
on which we predicted landscape permeability. We now expand on this knowledge
and use ISSF to develop a more detailed movement model of dispersing wild dogs.
We then use this model to simulate dispersers moving across the KAZA-TFCA. Based
on simulations, we compute heatmaps and identify potential dispersal hotspots.
We also showcase how network metrics relevant to landscape connectivity can be
computed. Our results show that a simulation-based approach yields several major
benefits over traditional connectivity modeling techniques. Most importantly,
simulations provide a more generic view on how connectivity emerges and to which
degree connectivity depends on the dispersal duration. In addtion, by generating
proper dispersal trajectories, network theory can be applied to calculate
network metrics that are pertinent to connectivity analysis. Finally, we put
forward additional opportunities using simulations that go beyond the scope of
this paper.

\section{Methods}
\subsection{Study Area (90\%)}
The study area was centered at -17\degree 13'9''S, 23\degree 56'4''E
(\Cref{StudyArea}a) and was represented by a rectangular bounding box that
stretched over 1.3 Mio. km\textsuperscript{2}, ecompassing the entire KAZA-TFCA
(\Cref{StudyArea}b). The KAZA-TFCA is the world's largest transboundary
conservation area and comprises parts of Angola, Botswana, Namibia, Zimbabwe,
and Zambia, covering a total of 520'000 km\textsuperscript{2}. Its landscape
varies regionally and ranges from savanna, to grassland and from dry to moist
woodland habitats. A dominant hydrogeographical feature in its center is the
flood-pulsing Okavango Delta. The wet season within the KAZA-TFCA lasts from
November to March and is out of phase with the main flooding of the Okavango
Delta which peaks between July and August \citep{McNutt.1996, Wolski.2017}.
While large portions within the KAZA-TFCA are designated national parks or other
protected areas, substantial human influence remains due to roads, agricultural
sites and settlements and villages.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_StudyArea.png}
    \caption{(a) The study area was confined by a bounding box encompassing the
    entire KAZA-TFCA and spanned parts of Angola, Namibia, Botswana, Zimbabwe,
    and Zambia. (b) The KAZA-TFCA represents the world's largest terrestrial
    conservation area and covers a total of 520'000 km\textsuperscript{2}. Its
    purpose is to re-establish connectivity between already-existing national
    parks (dark green) and other protected areas (light green). The dispersal
    data used in this study was collected on a free-ranging African wild dog
    population inhabiting the Moremi National Park in northern Botswana.}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{GPS Relocation Data (90\%)}
Between 2011 and 2019, we collected GPS relocation data on dispersing wild dogs
from a free-ranging wild dog population inhabiting the Moremi National Park in
northern Botswana. We selected potential dispersers based on age, pack size,
number of same‐sex siblings within the pack, and presence of unrelated
opposite-sex individuals in the pack \citep{McNutt.1996, Behr.2020}. We
immobilized selected individuals using a cocktail of Ketamine/Xylazine/Atropine
\citep{Osofsky.1996, Cozzi.2020} that was injected by dart, fired from a
CO\textsubscript{2}-pressurized gun (\textit{DAN-Inject, Denmark}). Immobilized
individuals were fitted with GPS/Satellite radio collars (\textit{Vertex Lite;
Vectronic Aerospace GmbH, Berlin}) that guaranteed automated drop-off through a
decomposable piece of cotton. Handling and collaring of all individuals was
supervised by a Botswana-registered wildlife veterinarian. All individuals
rejoined their pack within one hour after immobilization. 16 collared
individuals eventually dispersed, each in a separate same-sex dispersal
coalition (7 female and 9 male coalitions).

Because behavior during dispersal is more pertinent for assessing landscape
connectivity \citep{Elliot.2014, Abrahms.2017}, we discarded all data that was
collected during residency and only retained GPS data recorded during dispersal.
During dispersal, collars were programmed to record a GPS fix every 4 hours.
Collected relocations were regularly transmitted over the Iridium satellite
system, which allowed remote tracking of individuals, even if they left the main
study area and crossed international borders. In some instances, exact dispersal
dates were known from field observations. Otherwise, we distinguished between
residency and dispersal using the net-squared displacement metric. This metric
measures the squared Euclidean distance of a GPS relocation to a reference point
\citep{Borger.2012}, which in our case was set to the center of each
individual's natal home range. As such, dispersal was deemed to have started
when an individual left its natal home range and ended once individuals became
sedentary again. As previous research revealed similar behavior of females and
males during dispersal \citep{Woodroffe.2019, Cozzi.2020}, we did not
distinguish between sexes. After collection, we converted collected GPS
coordinates (n = 4'169) to steps, where each step represented the straight-line
distance traveled by and individual between two consecutive GPS relocations
\citep{Turchin.1998}. To ensure a regular sampling interval, we removed fixes
that were not successfully collected on the 4-hourly schedule.

\subsection{Covariates (90\%)}
We represented the physical landscape across the study area using a set of
habitat covariates including water-cover, distance to water, woodland-cover, and
shrub/grassland-cover. Because water cover greatly changes within and between
years in the Okavango Delta, we applied a remote sensing algorithm and generated
frequently updated water cover layers and corresponding distance to water layers
(see \citealp{Wolski.2017} and \citealp{Hofmann.2021}). Resulting layers thus
temporally aligned with our dispersal events. We furthermore computed a proxy
for human influence, depicting anthropogenic pressures stemming from
human-density, agricultural sites, and roads. All spatial layers were coarsened
or interpolated to a target resolution of 250 m by 250 m. Further details on the
sources and preparation of each habitat covariate are given in
\cite{Hofmann.2021}.

Besides habitat covariates, we also computed movement metrics that we used as
movement covariates in our models. Movement metrics were calculated for each
step and included the step length (\textsf{sl}), its natural logarithm
(\textsf{log(sl)}), and the cosine of the relative turning angle
(\textsf{cos(ta)}). Because wild dogs follow a diurnal activity pattern, we also
coded a binary variable (\textsf{LowActivity}) indicating whether a step was
realized during periods of low wild dog activity (17:00 to 07:00 local time) or
high wild dog activity (09:00 to 17:00 local time). Handling and manipulation of
all data, as well as all models and simulations were implemented with the
statistical software {\tt R}, version 3.6.6 \citep{R.2019}. Several helper
functions were written in {\tt C++} and imported into {\tt R} using the {\tt
Rcpp} package \citep{Eddelbuettel.2011, Eddelbuettel.2013}

\subsection{Movement Model (80\%)}
We used ISSFs to parametrize a mechanistic movement model for dispersing wild
dogs \citep{Avgar.2016}. To conduct iSSF analysis, we paired each observed step
with 24 random steps. An observed and its 24 random steps thus formed a stratum
and received a unique identifier. We generated random steps by sampling random
turning angles from a uniform distribution (\(-\pi, +\pi\)) and step lengths
from a gamma distirbution that was fitted to observed steps (scale = 6'308,
shape = 0.37). Along each step, we extracted and averaged spatial covariates
using the {\tt velox}. We also calculated the movement metrics \textsf{sl},
\textsf{log(sl)}, and \textsf{cos(ta)}. To facilitate model convergence, we
standardized all continuous covariates to a mean of zero and a standard
deviation of one. Since correlation among covariates was low (\(|r| > 0.6\);
\citealp{Latham.2011}), we retained all of them for modeling.

To contrast realized steps (scored 1) and random steps (scored 0), we assumed
that animals assigned a selection score \(w(x)\) of the exponential form to each
step. The selection score \(w(x)\) of each step depended on its associated
covariates (\(x_1, x_2, ..., x_n\)) and on the animal's relative selection
strengths (i.e. preferences) towards these covariates (\(\beta_1, \beta_2, ...,
\beta_n\)):

@@

The probability of a step being realized was then contingent on the step's
selection score, as well as on the selection scores of all other step in the
same stratum:

@@

To estimate the preferences of interest, we ran conditional logistic regression
models in the r-package {\tt glmmTMB}. To handle multiple individuals, we
applied the mixed effects method developed by \citep{Muff.2020}, which allows to
fit random slopes in addition to random intercepts. Thus, we treated animal IDs
as random effect and modeled random slopes for each covariate. We fixed the
random intercept variance at an arbitrary high value of 10\textsuperscript{6} to
make use of the  ``poission''-trick \citep{Muff.2020}.

The movement model was based on the habitat selection model for dispersing wild
dogs presented in \cite{Hofmann.2021}. In the original model, no interactions
between the habitat and movement covariates were considered, so we slightly
expanded this base model by proposing interactions between movement and habitat
covariates. More specifically, we started with the base model and incrementally
increased model complexity by adding all possible two-way interactions between
habitat covariates and movement covariates. For instance, for the covariate
\textsf{water}, we proposed the interactions \textsf{Water:log(sl)},
\textsf{Water:log(sl)}, and \textsf{Water:cos(ta)}. Besides those combinations,
we also proposed the interactions \textsf{sl:cos(ta)} and
\textsf{log(sl):cos(ta)} to account for a correlation between turning angles and
step lengths, as well as the interactions \textsf{sl:LowActivity} and
\textsf{log(sl):LowActivity} to account for the fact that step lengths may
differ due to wild dogs' diurnal activity pattern. To compare competing models
and assess the most parsimonious movement model, we ran stepwise forward model
selection based on Akaike's Information Criterion (AIC, \citealp{Burnham.2002}).
